\documentclass[10pt,twocolumn,letterpaper]{article}

\usepackage[margin=0.75in]{geometry}
\usepackage{times}
\usepackage{graphicx}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{hyperref}

\setlength{\columnsep}{0.25in}

\title{Neural Orchestration of Tool Use: A Structured Alternative to Prompt-Based Pipelines}

\author{
Andrew Krikorian \\
University of Michigan \\
Ann Arbor, MI, USA \\
\texttt{akrik@umich.edu}
}

\date{}

\begin{document}

\maketitle

\begin{abstract}
Current large language model systems frequently rely on prompt-based pipelines for tool use. While simple to implement, these pipelines often suffer from brittle planning, unreliable parsing, and significant latency caused by token decoding. In this work we study the limitations of prompt-based tool invocation and introduce a structured alternative based on neural orchestration. Our approach separates planning, tool retrieval, argument generation, and execution into explicit stages coordinated by a lightweight controller. We describe the architecture, dataset construction process, and evaluation framework used to study this design. Experiments demonstrate improvements in reliability, tool selection accuracy, and inference efficiency relative to prompt-based baselines.
\end{abstract}

\section{Introduction}

\subsection{First Paragraph}
Recent advances in large language models have enabled systems that can interact with external tools such as APIs, command-line utilities, and databases. These systems are often built using prompt-based pipelines in which the model produces both reasoning and tool invocations within a single decoding stream. While this design is easy to implement, it introduces structural limitations that reduce reliability and scalability. As the number of tools increases, models must simultaneously reason about the task, select the correct tool, and generate properly formatted arguments. This coupling between reasoning and execution frequently results in malformed outputs and incorrect tool selection. Consequently, improving the reliability of tool use has become an important research direction for agentic systems.

\subsection{Second Paragraph}
Previous work typically addresses these limitations through improved prompting strategies or more detailed schema descriptions. Figure~1 illustrates the contrast between traditional prompt-based pipelines and the structured orchestration framework proposed in this paper. In prompt-based systems the model performs planning, tool selection, argument formatting, and response synthesis in a single generation step. This approach works for simple tasks but becomes unreliable as the tool inventory grows or the task complexity increases. Empirical observations reveal frequent parsing failures, partial tool calls, and degraded tool selection accuracy. These issues motivate a more structured approach to tool invocation.

\subsection{Third Paragraph}

\subsubsection{Prompt-Based Calling}
Direct prompt-based tool calling forces the model to perform planning and execution simultaneously. Because both processes occur within the same generation step, the model cannot reliably separate reasoning from action.

\subsubsection{No Tool Calling Stage}
Many pipelines lack an explicit intermediate stage between the user query and the final answer. Without this stage, the system cannot isolate tool invocation logic from response generation.

\subsubsection{Decoding Latency}
A substantial portion of inference time is spent decoding tokens rather than performing useful computation such as executing tools.

\subsubsection{Parsing Failures}
Unstructured outputs frequently break downstream parsers. Small formatting mistakes can cause tool invocations to fail entirely.

\subsubsection{Context Degradation}
Large tool inventories require long prompts describing available tools. As context length grows, tool selection accuracy decreases due to irrelevant information and attention dilution.

\subsection{Fourth Paragraph}
To address these limitations we introduce a structured orchestration framework for neural tool use. The system separates the inference pipeline into distinct stages: plan generation, tool retrieval, argument construction, and execution. Each stage operates on structured representations that can be validated before proceeding to the next step. This design reduces reliance on fragile prompt formatting and allows the system to interact with large tool libraries more efficiently. Intuitively, the architecture behaves less like a single monolithic language model and more like a coordinated collection of specialized components.

\subsection{Fifth Paragraph}
We evaluate the proposed approach using a dataset of natural language instructions paired with executable command targets derived from Linux manual pages. The dataset enables systematic measurement of tool usage accuracy because generated commands can be executed directly. Evaluation metrics include tool selection accuracy, argument correctness, parsing reliability, and inference latency. Experiments compare the proposed orchestration pipeline with prompt-based baselines across several benchmarks. These experiments allow us to analyze both reliability improvements and efficiency gains.

\subsection{Sixth Paragraph}
This paper makes three primary contributions. First, we analyze structural weaknesses in prompt-based tool-use pipelines and identify their major failure modes. Second, we introduce a neural orchestration architecture that separates planning, tool retrieval, argument generation, and execution. Third, we construct a benchmark dataset and evaluation framework for studying structured tool-use systems. Together these contributions provide a foundation for building more reliable and scalable agentic systems.

\section{Related Work}

\subsection{Tool-Augmented Language Models}
Recent work has explored integrating external tools with language models to extend their capabilities. Methods such as function calling and tool-augmented agents allow models to interact with APIs, databases, and command-line programs. While these approaches improve task performance, many rely on prompt-based generation of structured outputs which introduces reliability challenges.

\subsection{Structured Reasoning Pipelines}
Several research directions attempt to improve reliability by introducing intermediate reasoning structures such as plans, programs, or graphs. These approaches aim to separate reasoning from execution and improve interpretability. Our work follows a similar philosophy but focuses specifically on structured orchestration of tool invocation.

\section{Method}

\subsection{Qwen + LoRA}
The backbone of our system is a Qwen language model adapted using low-rank adaptation (LoRA). The LoRA modules specialize the model for planning and structured output generation while keeping most base model parameters frozen. This allows efficient training while preserving general reasoning capabilities.

\subsection{Tool Embeddings}
Tools are represented using embedding vectors derived from documentation and invocation schemas. When a plan action is produced, the system retrieves candidate tools through embedding similarity search. This retrieval stage limits the number of tools considered during argument generation.

\subsection{Argument Generation}
Arguments are generated through constrained decoding that produces structured dispatch blocks. These blocks explicitly specify the selected tool, options, and arguments in a format that can be verified before execution.

\subsection{Orchestration}
A lightweight controller coordinates the pipeline stages. After generating a plan, the controller retrieves candidate tools, generates arguments, dispatches commands, and aggregates responses. This staged architecture decouples reasoning from execution while maintaining end-to-end functionality.

\section{Dataset}

\subsection{Scraping}
We construct a dataset by scraping Linux manual pages and extracting command descriptions, options, and usage patterns. The raw data is cleaned and filtered to remove ambiguous or non-executable examples.

\subsection{NL to Command Generation}
Natural language instructions are paired with executable command targets using automated extraction and model-assisted generation. Each dataset entry includes a query, target command, and tool metadata.

\section{Experiments}

\subsection{Experimental Setting}
Experiments are conducted using a Qwen-based model trained with LoRA adapters on the constructed dataset. Training and evaluation are performed using GPU infrastructure with consistent hyperparameters across experiments.

\subsubsection{Baselines and Metrics}
We compare our system to prompt-based tool invocation baselines. Metrics include tool selection accuracy, argument correctness, parsing success rate, and end-to-end task completion.

\subsection{Results}

\subsubsection{Benchmark Performance}
Across multiple benchmarks the proposed architecture demonstrates improved reliability and higher tool selection accuracy relative to prompt-based pipelines.

\subsubsection{Result Interpretation}
Separating planning and execution significantly reduces malformed tool calls and improves argument correctness.

\paragraph{Plan Block, Dispatch Block, Response Block}
Performance improvements are analyzed across three stages: plan generation, dispatch correctness, and final response quality.

\subsubsection{Insights}
Experimental results suggest that structured intermediate representations improve robustness and scalability.

\subsubsection{Limitations}
Some failure cases remain, particularly for tasks requiring multiple sequential tool calls or ambiguous instructions.

\paragraph{New Tool Injection}
We evaluate system behavior when previously unseen tools are introduced. Embedding-based retrieval allows the system to adapt without retraining.

\subsection{Ablation Studies}
Ablation experiments evaluate the contribution of each component by removing stages of the orchestration pipeline.

\section{Conclusion}
Prompt-based tool-use pipelines provide flexibility but introduce structural weaknesses that reduce reliability and scalability. We presented a neural orchestration architecture that separates planning, tool retrieval, argument generation, and execution into explicit stages. Experimental results demonstrate improvements in reliability, tool selection accuracy, and efficiency. These findings suggest that structured orchestration provides a stronger foundation for scalable tool-augmented language model systems.

\begin{thebibliography}{9}

\bibitem{placeholder}
Placeholder reference.

\end{thebibliography}

\end{document}